---
title: "M5120 Project"
subtitle: "Weather Prediction"
author: "Aryna Pethshok, Jiri Hofirek, Ondrej Svihnos"
date: ""
format: 
  html: 
    page-layout: full
    embed-resources: true
    toc: true
    toc-location: left
    number-sections: true
    code-tools:
      source: true
---
# Exploration analysis

```{r}
library(dplyr)
library(ggplot2)
library(tidyr)
library(ggcorrplot)
library(GGally)
library(lubridate)

weather <- read.csv("weatherHistory.csv")
weather <- select(weather, !Loud.Cover)
names(weather)
```


```{r}
# corr <- weather %>%
#   select(Temperature, Apparent.Temperature, Humidity, Wind.Speed, Wind.Bearing, Visibility, Pressure) %>%
#   cor()
weather %>% 
  filter(!(weather$Pressure==0)) %>%
  select(Temperature, Humidity, Pressure) %>%
  ggpairs(
    lower = list(continuous = wrap("points", color = "deepskyblue", alpha = 0.3)),
    diag = list(continuous = wrap("barDiag", fill = "tomato", color = "black"))
    )

#ggcorrplot(corr, type = 'upper', lab = T)

##density of variables (už je v ggpairs)
# weather %>%
#   ggplot(aes(x = Temperature, y = after_stat(density))) +
#   geom_histogram(fill = "Orange", alpha = 0.5,
#                  color = 'Orange',
#                  bins = 30) +
#   labs(x = "Temperature [C]", y = "Density") +
#   theme_bw()
# 
# weather %>%
#   ggplot(aes(x = Apparent.Temperature, y = after_stat(density))) +
#   geom_histogram(fill = "Orange", alpha = 0.5,
#                  color = 'Orange',
#                  bins = 30) +
#   labs(x = "Apparent temperature [C]", y = "Density") +
#   theme_bw()
# 
# weather %>%
#   ggplot(aes(x = Wind.Bearing, y = after_stat(density))) +
#   geom_histogram(fill = "Blue", alpha = 0.5,
#                  color = 'Blue',
#                  bins = 30) +
#   labs(x = "Wind direction [angles]", y = "Density") +
#   theme_bw()
# 
# weather %>%
#   ggplot(aes(x = Visibility, y = after_stat(density))) +
#   geom_histogram(fill = "Orange", alpha = 0.5,
#                  color = 'Orange',
#                  bins = 30) +
#   labs(x = "Visibility [km]", y = "Density") +
#   theme_bw()
# 
# weather %>%
#   ggplot(aes(x = Temperature, y = after_stat(density))) +
#   geom_histogram(fill = "Orange", alpha = 0.5,
#                  color = 'Orange',
#                  bins = 30) +
#   labs(x = "Temperature [C]", y = "Density") +
#   theme_bw()

##corelation of variables
# weather %>%
#   filter(Precip.Type == "null") %>%
#   ggplot(aes(x = Humidity, y = Temperature, color=Precip.Type)) +
#   geom_point(alpha=0.5)
# 
# weather %>%
#   ggplot(aes(x=Temperature, y=Apparent.Temperature, color=Wind.Speed, fill = Wind.Bearing)) + 
#   geom_point(alpha=0.5)
# 
# weather %>%
#   ggplot(aes(x=Wind.Bearing, y=Wind.Speed, color=Pressure))+
#   geom_point(alpha=0.5)
# 
# weather %>%
#   ggplot(aes(x=Formatted.Date, y=Temperature))+
#   geom_point(alpha=0.1)+
#   theme_bw()
# 
# weather %>%
#   filter(Precip.Type == "null") %>%
#   ggplot(aes(x=Visibility, y=Humidity, color=Precip.Type))+
#   geom_point(alpha=0.5)


```



#  Preprocessing
## Create Factor variable, Format Data Time,  drop Summary variables.
```{r}
str(weather)
```


```{r}
unique(weather$Summary)
unique(weather$Precip.Type)
```
```{r}
#Drop Summary columns
weather <- weather %>%
  select(-Summary, -Daily.Summary)
```

```{r}
# Convert categorical columns to factors
weather <- weather %>%
  mutate(Precip.Type = factor(Precip.Type)) %>%
  mutate(Formatted.Date = lubridate::ymd_hms(Formatted.Date, tz = "UTC"))
```

## Zero values

```{r}
zero_summary <- weather %>%
  select(where(is.numeric)) %>%
  summarise(across(everything(), ~ sum(. == 0, na.rm = TRUE))) %>%
  pivot_longer(cols = everything(),
               names_to = "Column",
               values_to = "Zero_Count") %>%
  mutate(
    Total_Values = sapply(weather[Column], function(x) sum(!is.na(x))),
    Zero_Percent = (Zero_Count / Total_Values) * 100
  )
zero_summary
```

```{r}
ggplot(zero_summary, aes(x = reorder(Column, -Zero_Count), y = Zero_Count)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  geom_text(aes(label = Zero_Count), vjust = -0.3) +
  labs(
    title = "Number of Zero Values",
    x = "Column",
    y = "Zero Count"
  ) +
  theme_minimal()
```

Pressure can be considered zero in a near-perfect vacuum, such as outer space. On Earth, this is not a realistic scenario. 
Therefore we will replace 0 values for Pressure with the calculated pressure, which depends on a Temperature.

### Zero values for Pressure

```{r}
P0 <- 101325         
h  <- 80            #Szeged is roughly 80 meters above sea level
g  <- 9.807          
M  <- 0.0289644     
R  <- 8.31447        

weather <- weather %>%
  mutate(
    Pressure = if_else(
      Pressure == 0,
      P0 * exp(-g * M * h / (R * (Temperature + 273.15))),
      Pressure
    )
  )
```

### Zero values for Wind.Speed and Wind.Bearing
Logically, if Wind.Speed is Zero, there is no meaningful wind direction.Therefore we will first analyze data for 0 Values Speed.
```{r}
wind0_summary <- weather %>%
  filter(Wind.Bearing == 0) %>% 
  mutate(
    Category = ifelse(
      Wind.Speed == 0,
      "Bearing 0, Wind = 0",
      "Bearing 0, Wind > 0"   
    )
  ) %>%
  group_by(Category) %>%
  summarise(Count = n(), .groups = "drop")

wind0_summary
```

```{r}
weather <- weather %>%
  mutate(
    Direction = ifelse(
      Wind.Speed == 0,
      "No direction",
      as.character(cut(
        Wind.Bearing,
        breaks = seq(0, 360, by = 45),
        labels = c("N", "NE", "E", "SE", "S", "SW", "W", "NW"),
        include.lowest = TRUE
      ))
    ),
    Direction = factor(Direction, 
                       levels = c("No direction", "N", "NE", "E", "SE", "S", "SW", "W", "NW"))
  )

wind_summary <- weather %>%
  group_by(Direction) %>%
  summarise(
    Mean_Wind_Speed = mean(Wind.Speed, na.rm = TRUE),
    Count = n(),
    .groups = "drop"
  )
wind_summary

ggplot(wind_summary, aes(x = Direction, y = Mean_Wind_Speed, fill = Mean_Wind_Speed)) +
  geom_col(color = "black", size = 0.2) +
  coord_polar(start = -pi/8) +
  scale_fill_gradient(low = "lightblue", high = "darkblue") +
  labs(
    title = "Mean Wind Speed by Direction",
    fill = "Wind Speed"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(size = 10),
    legend.position = "right"
  )

```

## Remove duplicetes.
```{r}
weather %>% 
  duplicated() %>% 
  sum()
```

```{r}
weather <- weather %>% 
  filter(!duplicated(weather))
```

## Outliers

```{r}
numeric_cols <- sapply(weather, is.numeric)
for(col_name in names(weather)[numeric_cols]) {
  boxplot(weather[[col_name]],
          main = paste(col_name),
          ylab = col_name,
          col = "lightblue",
          border = "darkblue")

}
```
Temperature, Apparent Temperature, Humidity, Wind Speed and Pressure -  those columns have some outliers values.

```{r}
# Temperature
Q1_temperature <- quantile(weather$Temperature, 0.25, na.rm = TRUE)
Q3_temperature <- quantile(weather$Temperature, 0.75, na.rm = TRUE)
IQR_temperature <- Q3_temperature - Q1_temperature

lower_temperature <- Q1_temperature - 1.5 * IQR_temperature
upper_temperature <- Q3_temperature + 1.5 * IQR_temperature

# Apparent Temperature
lower_app_temperature <- quantile(weather$Apparent.Temperature, 0.05, na.rm = TRUE)
upper_app_temperature <- quantile(weather$Apparent.Temperature, 0.95, na.rm = TRUE)

# Humidity
Q1_humidity <- quantile(weather$Humidity, 0.25, na.rm = TRUE)
Q3_humidity <- quantile(weather$Humidity, 0.75, na.rm = TRUE)
IQR_humidity <- Q3_humidity - Q1_humidity

lower_humidity <- Q1_humidity - 1.5 * IQR_humidity
upper_humidity <- Q3_humidity + 1.5 * IQR_humidity

# Wind Speed
lower_wind <- quantile(weather$Wind.Speed, 0.05, na.rm = TRUE)
upper_wind <- quantile(weather$Wind.Speed, 0.95, na.rm = TRUE)

# Pressure
lower_pressure <- quantile(weather$Pressure, 0.05, na.rm = TRUE)
upper_pressure <- quantile(weather$Pressure, 0.95, na.rm = TRUE)

# Replacing outliers with lower and upper bounds
weather <- weather %>%
  mutate(
    Temperature = ifelse(Temperature < lower_temperature, lower_temperature,
                         ifelse(Temperature > upper_temperature, upper_temperature, Temperature)),
    
    Apparent.Temperature = ifelse(Apparent.Temperature < lower_app_temperature, lower_app_temperature,
                                  ifelse(Apparent.Temperature > upper_app_temperature, upper_app_temperature, Apparent.Temperature)),
    
    Humidity = ifelse(Humidity < lower_humidity, lower_humidity,
                      ifelse(Humidity > upper_humidity, upper_humidity, Humidity)),
    
    Wind.Speed = ifelse(Wind.Speed < lower_wind, lower_wind,
                        ifelse(Wind.Speed > upper_wind, upper_wind, Wind.Speed)),
    
    Pressure = ifelse(Pressure < lower_pressure, lower_pressure,
                      ifelse(Pressure > upper_pressure, upper_pressure, Pressure))
  )

numeric_cols <- sapply(weather, is.numeric)
for(col_name in names(weather)[numeric_cols]) {
  boxplot(weather[[col_name]],
          main = paste(col_name),
          ylab = col_name,
          col = "lightblue",
          border = "darkblue")
}

```

## Creating new variables for Temperature, Humidity, Pressure hour before. 

```{r}
#First row value is replaced by the second value, to avoid NA values.
weather <- weather %>%
  arrange(Formatted.Date) %>%
  mutate(
    Temperature_prev_hour = coalesce(lag(Temperature, 1), first(Temperature)),
    Humidity_prev_hour    = coalesce(lag(Humidity, 1), first(Humidity)),
    Pressure_prev_hour    = coalesce(lag(Pressure, 1), first(Pressure))
  )
```





